---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---


You will need to leverage a number of R packages when working with this Notebook. You can add these packages by entering the following line in your Console:
install.packages(c("tidyverse","leaflet","lubridate","esquisse","viridis","sf"))

```{r}
library(tidyverse)
library(leaflet)
library(lubridate)
library(viridis)
```

## Data Loading

```{r}
affordable_housing_sf <- read.csv("https://data.sfgov.org/resource/aaxw-2cb8.csv")

```

## Data Cleaning

```{r}
#Convert NAs
is.na(affordable_housing_sf) <- affordable_housing_sf == ""

#Convert dates
affordable_housing_sf$issuance_of_notice_to_proceed <- ymd_hms(affordable_housing_sf$issuance_of_notice_to_proceed)
affordable_housing_sf$issuance_of_building_permit <- ymd_hms(affordable_housing_sf$issuance_of_building_permit)
affordable_housing_sf$issuance_of_first_construction_document <- ymd_hms(affordable_housing_sf$issuance_of_first_construction_document)
affordable_housing_sf$estimated_actual_construction_start_date <- ymd_hms(affordable_housing_sf$estimated_actual_construction_start_date)
affordable_housing_sf$estimated_construction_completion <- ymd_hms(affordable_housing_sf$estimated_construction_completion)
affordable_housing_sf$entitlement_approval <- ymd_hms(affordable_housing_sf$entitlement_approval)
```

Run: 
esquisse::esquisser()

## Descriptive Data Analysis Plots

//Write your first question here.

```{r}
#Copy your first plot here
```

//Write your second question here.

```{r}
#Copy your second plot here
```

//Write your third question here.

```{r}
#Copy your third plot here
```





